A Quasi-Genuine FIFO Total Order Multicast Primitive

### 2011-07-19 trunk/textos/optpaxos/draftv3.tex

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:mcastbarpending}
Once a message has been \cms{}, it will be inserted into the $barPending$ list of the correct processes to which such message has been addressed.
\end{lems}

\begin{proof}
In Algorithm \ref{algorithm:deliveryminimal}, whenever a process in a group $G$ \cms{}s a message, it is first \rmc{} to all processes in $G$ (l. \ref{algline:rmlocal}). As the \rmc{} is reliable, every correct process from $G$ \rmd{}s $m$. Once $m$ is \rmd{}ed by each process $p \in G$, it is inserted into the $propPending$ list of that process, to be included in the next proposal from $p$. As every process from $G$ received $m$ and enqueued it for being proposed, and since $m$ is only removed from $propPending$ when decided (l. \ref{algline:removedecided}), all processes will proposed it at some point and $m$ will eventually be decided. At this point, if $G \in m.dst$, every correct process of $G$ will insert $m$ into its $barPending$ list (l. \ref{algline:insbp2}). Besides, each correct process in $G$ \rmc{}s $m$ to all other destination groups that $m$ might be adressed to.

@@

\begin{proof}Every process $q$ keeps executing Algorithm \ref{algorithm:nullperiodic}, thus sending ever increasing barrier values to the processes in $receiversFrom(group(q))$. From the given definitions, it is obvious that if $G \in receiversFrom(G')$, then $G' \in sendersTo(G)$. Besides, once decided and inserted into the $barPending$ list, the timestamp of the messages no longer can change. This means that, at some point, all of the barrier values received by $p$ will be greater than $m.ts$.
\end{proof}

@@

Assume, by way of contradiction, that $p$ received a message $m':m'.ts < m.ts$ from some process $p'' \in G'$ after having received $m$ from $p'$. As $m'.src = m.src$, this means that $p$ either will send $m'$ afterwards, contradicting the fact that the processes in $G'$ agree on each batch of messages decided and send them to $G$ in the same FIFO order, or that $p'$ will not send $m'$ at all, contradicting the fact the the \rmc{} primitive is reliable. Therefore, the assumption is false and the lemma is valid.
\end{proof}

@@

\begin{proof}
Immediate from the algorithm, considering that a message is never removed from the $delivered$ list, and that it is removed from $barPending$ (l. \ref{algline:rmfrombp}) only to be inserted into $delivered$ (l. \ref{algline:addtodelivered}).
\end{proof}

@@

\begin{proof}Let $p$ be a process from the group $G$ which sent $m$. Before inserting $m$ into the $decided$ list (l. \ref{algline:addtodecided} of Algorithm \ref{algorithm:deliveryminimal}), $p$ checks whether there is already some other message with an equal or lower timestamp in such list. If that is the case, the timestamp of $m$ is changed to a value greater than that of any other message already within $decided$. Only then, if $G$ is a destination of $m$, $p$ inserts $m$ into its $barPending$ list (l. \ref{algline:insbp2}). Therefore, for processes from the group from which $m$ was sent, it is inserted in the $barPending$ list always in ascending order of timestamp. Besides, each process from $G$ \rmc{}s $m$ to its other destination groups in this same order (l. \ref{algline:rmothers}).

Regarding groups other than the source of the message, let $p'$ be any process that is a destination of $m$, but does not belong to the group $G$ from where $m$ came. The process $p'$ inserts $m$ into its $barPending$ list in line \ref{algline:insbp1}. Considering that every process in a group sends that group's messages to other groups in the same order in which they are decided, that once a group has received some message from another group, any message prior to that one from the same group has also already been received (from Lemma \ref{lemma:groupfifo}), and that no message is inserted twice in the $barPending$ list of any process (from line \ref{algline:notbptwice} and Lemma \ref{lemma:alwaysbpdel}), we can infer that, once $p'$ inserts $m$ into its $barPending$ list, no message $m' : m'.ts < m.ts$ from $G$ will be inserted into the same $barPending$ list afterwards.

\end{proof}

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:oncebpwilldel}
Once a message $m$ has been inserted into the $barPending$ list of a process $p$, it will eventually be removed from this list and, if $m \neq null$ it will be delivered by $p$.
\end{lems}

\begin{proof}
Eventually, $p$ will have received a barrier from every group in $sendersTo(group(p))$ with a value greater than the timestamp of $m$ (from Lemma \ref{lemma:barrierperiodic}). Once this happens, as any barrier is also a message, no more messages with a timestamp lower than that of $m$ will be received by $p$ (from Lemma \ref{lemma:oncedecidednotsmaller}). Let $barReady$ be the set of $m$ plus all messages from $barPending$ whose timestamps are lower than that of $m$, that is, $barReady = \{m\} \cup \{m' : m' \in barPending \wedge m'.ts < m.ts\}$.

It is important to note that, from Lemma \ref{lemma:oncedecidednotsmaller}, no more messages will be included in this set, meaning that any other message in $barPending$ will have a timestamp greater than that of any message in $barReady$. Besides, the barrier received for $m$ also apply to all the other messages in this set, since their timestamps are lower than that of $m$. We can, therefore, prove that all these messages will eventually be delivered by induction on the position $i$ of each message $m_i$ in $barReady$, in ascending order of timestamps.\\
\\
\noindent Base case ($i=1$): Let $m_1$ be the first message in $barReady$. It means that $\nexists m \in barPending : m.ts < m_1.ts$. The message $m_1$ has the lowest timestamp in the list, all the necessary barriers have been received and no more messages with lower timestamps will arrive at $p$. Therefore, $m_1$ satisfies all conditions from line \ref{algline:inbarpending} of Algorithm \ref{algorithm:deliveryminimal} and will be removed from $barPending$. If it is different from $null$, it will also be delivered.\\
\\
\noindent Induction step: Suppose that $m_{i}$ will eventually be removed from $barPending$. Once this happens, it means that $m_{i}$ was the first message in the list. As soon as $m_{i}$ is removed, $m_{i+1}$ will be the new first one. Then, as $m_{i+1}$ will have the lowest timestamp in $barReady$ -- and, therefore, in $barPending$ --, it will satisfy both conditions of line \ref{algline:inbarpending} from Algorithm \ref{algorithm:deliveryminimal}), being removed from $barPending$ -- and delivered, if it is different from $null$. \end{proof}

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props}
Once a process \cms{}s a message, then one of the correct processes that is a destination of that message eventually delivers it.
\end{props}

\begin{proof}
Immediate from Lemma \ref{lemma:mcastbarpending}, Lemma \ref{lemma:oncebpwilldel}, and from the assumption that the application will not \cms{} a $null$ message -- not in the \cms{} protocol level at least, since the application might need to \cms{} an empty message, which in the protocol level would be handled as a meaningful one.
\end{proof}

@@

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props}
If a message $m$ was delivered to process $p$ of group $G$, then it has (1) been \cms{} before, (2) $G \in m.dst$ and (3) it has not been \cons{}ed by $p$ before.
\end{props}

@@

\subsection{Uniform Total Order}

@@

>>>>

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:mcastbarpending}
Once a message has been \cms{}, it will be inserted into the $barPending$ list of all the correct processes to which such message has been addressed.
\end{lems}

\begin{proof}
In Algorithm \ref{algorithm:deliveryminimal}, whenever a process in a group $G$ \cms{}s a message, it is first \rmc{} to all processes in $G$ (l. \ref{algline:rmlocal}). From the uniform agreement property of the \rmc{} primitive, every correct process from $G$ \rmd{}s $m$. Once $m$ is \rmd{}ed by each process $p \in G$, it is inserted into the $propPending$ list of that process, to be included in the next proposal from $p$. As every process from $G$ received $m$ and enqueued it for being proposed, and since $m$ is only removed from $propPending$ when decided (l. \ref{algline:removedecided}), every processes of $G$ will propose it at some point and $m$ will eventually be decided. At this point, if $G \in m.dst$, every correct process of $G$ will insert $m$ into its $barPending$ list (l. \ref{algline:insbp2}). Besides, each correct process in $G$ \rmc{}s $m$ to all other destination groups that $m$ might be adressed to.

@@

\begin{proof}Every process $p'$ keeps executing Algorithm \ref{algorithm:nullperiodic}, thus sending ever-increasing barrier values to all processes in $receiversFrom(group(p'))$. Also, if $G \in receiversFrom(G')$, then $G' \in sendersTo(G)$. This means that, at some point, all correct processes of each group in $sendersTo(G)$ will have inserted a message -- be it $null$ or not -- in their $propPending$ lists, with a timestamp greater than, or equal to, that of $m$. As this is done by every process in each of the groups in $sendersTo(G)$, and a message is removed from $propPending$ only when decided, then eventually $null$ will be decided and \rmc{} to $G$. From the uniform validity property of \rmc{}, such messages will be delivered by $p$, which will insert them into its $barPending$ list. From the algorithm, the timestamps of the messages are no longer changeable after they have been inserted into $barPending$ so, eventually, $p$ will have in such list a barrier with a value greater than or equal to $m.ts$ from each group in $sendersTo(G)$. 
\end{proof}

@@

Assume, by way of contradiction, that $p$ received a message $m':m'.ts < m.ts$ from some process $p'' \in G'$ after having received $m$ from $p'$. As $m'.src = m.src$, this means that $p'$ either will send $m'$ afterwards, contradicting the fact that the processes in $G'$ agree on each batch of messages decided and send them to $G$ in the same FIFO order, or that $p'$ will not send $m'$ at all, contradicting the uniform validity property of the \rmc{} primitive. Therefore, the assumption is false and the lemma is valid.
\end{proof}

@@

\begin{proof}
Immediate from the algorithm, considering that a message is never removed from the $delivered$ list, and that it is removed from $barPending$ (l. \ref{algline:rmfrombp}) right before being inserted into $delivered$ (l. \ref{algline:addtodelivered}).
\end{proof}

@@

\begin{proof}
Let $p$ be a process from the group $G$ which sent $m$. Before inserting $m$ into the $decided$ list (l. \ref{algline:addtodecided} of Algorithm \ref{algorithm:deliveryminimal}), $p$ checks whether there is already some other message with an equal or lower timestamp in such list. If that is the case, the timestamp of $m$ is changed to a value greater than that of any other message already within $decided$. Only then $m$ may be inserted into the $barPending$ list of $p$ (l. \ref{algline:insbp2}). Therefore, $p$ inserts messages from other processes in $group(p)$ into its $barPending$ list in ascending order of timestamps. Besides, each process from $G$ \rmc{}s $m$ to the other destinations of $m$ in this same order (l. \ref{algline:rmothers}).

Regarding groups other than the source of the message, let $p'$ be any process that is a destination of $m$, such that $p' \in G' \wedge G' \neq G$. The process $p'$ inserts $m$ into its $barPending$ list in line \ref{algline:insbp1}. Considering that every process in a group sends that group's messages to other groups in the same order in which they are decided, that once a group has received some message from another group, any message prior to that one from the same group has also already been received (from Lemma \ref{lemma:groupfifo}), and that no message is inserted twice in the $barPending$ list of any process (from line \ref{algline:notbptwice} and Lemma \ref{lemma:alwaysbpdel}), we can infer that, once $p'$ inserts $m$ into its $barPending$ list, no message $m' : m'.ts < m.ts$ from $G$ will be inserted into the same $barPending$ list afterwards.
\end{proof}

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:oncebpwilldel}
Once a message $m$ has been inserted into the $barPending$ list of a process $p$, it will eventually be removed from such list and, if $m$ is different from $null$, it will be delivered by $p$.
\end{lems}

\begin{proof}
Eventually, $p$ will have received a barrier from every group in $sendersTo(group(p))$ with a value greater than the timestamp of $m$ (from Lemma \ref{lemma:barrierperiodic}). Once this happens, as any barrier is also a message, no more messages with a timestamp lower than that of $m$ will be inserted into the $barPending$ list of $p$ (from Lemma \ref{lemma:oncedecidednotsmaller}). Let $barReady$ be the set of $m$ plus all messages from $barPending$ whose timestamps are lower than that of $m$, that is, $barReady = \{m\} \cup \{m' : m' \in barPending \wedge m'.ts < m.ts\}$.

We can infer that no more messages will be included in this set, meaning that any other message that might be later inserted into $barPending$ will have a timestamp greater than that of any message in $barReady$. Besides, the barrier received for $m$ also apply to all the other messages in this set, since their timestamps are lower than that of $m$. We can, therefore, prove that all these messages will eventually be delivered by induction on the position $i$ of each message $m_i$ in $barReady$, in ascending order of timestamps.\\
\\
\noindent Base case ($i=1$): Let $m_1$ be the first message in $barReady$. It means that $\nexists m \in barPending : m.ts < m_1.ts$. The message $m_1$ has the lowest timestamp in the list and all the necessary barriers have been received already. Therefore, $m_1$ satisfies all conditions from line \ref{algline:inbarpending} of Algorithm \ref{algorithm:deliveryminimal} and will be removed from $barPending$. If it is different from $null$, it will also be delivered.\\
\\
\noindent Induction step: Suppose that $m_{i}$ will eventually be removed from $barPending$. Once this happens, it means that $m_{i}$ was the first message in the list. Since no more messages have been inserted into $barReady$, as soon as $m_{i}$ is removed, $m_{i+1}$ will be the new first one. Then, as $m_{i+1}$ will have the lowest timestamp in $barReady$, it will also have the lowest timestamp of $barReady$. As all barriers necessary for $m$ have already been received and $m_{i+1}.ts \leq m.ts$, it will satisfy both conditions of line \ref{algline:inbarpending} from Algorithm \ref{algorithm:deliveryminimal}. Thus, $m_{i+1}$ will be removed from $barPending$ -- and delivered, if it is different from $null$. \end{proof}

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props} \label{props:validity}
Once a process \cms{}s a message, then all correct processes that are destinations of that message eventually deliver it.
\end{props}

\begin{proof}
Immediate from Lemma \ref{lemma:mcastbarpending}, Lemma \ref{lemma:oncebpwilldel}, and from the assumption that the application will not \cms{} a $null$ message -- not in the \cms{} protocol level at least; if the application needs to \cms{} an empty message, in the protocol level it would not be handled as a $null$ message.
\end{proof}

@@

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props} \label{props:integrity}
If a message $m$ was delivered to process $p$ of group $G$, then it has (1) been \cms{} before, (2) $G \in m.dst$ and (3) it has not been \cons{}ed by $p$ before.
\end{props}

@@

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props} \label{props:agreement}
If a correct process which is a destination of a message $m$ delivers it, then all correct processes that are also destinations of $m$ deliver it as well.
\end{props}

\begin{proof}
Immediate from Proposition \ref{props:integrity}, Lemma \ref{lemma:mcastbarpending} and Lemma \ref{lemma:oncebpwilldel}.
\end{proof}

\subsection{Atomic Order}

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:agreetimestamps}
If two correct processes are both destinations of a message $m$, then they agree on its final timestamp $m.ts$.
\end{lems}

\begin{proof}
Let $p$ be the process that is \cms{}ing $m$. We can prove this lemma by demonstrating that any other process $p'$ will agree with $p$ on the timestamp of $m$. Let us first consider the case of $p'$ belonging to the same group as $p$. We can prove by induction on the identifier $k$ of each consensus instance within $group(p)$:\\

\noindent Base case ($k=1$): As this is the first consensus instance of the group, it means that the $decided = \emptyset$. From the agreement property of consensus, we know that $p$ and $p'$ decided the same contents for $msgSet$ \mbox{(l. \ref{algline:decide})}. Each message included in such agreed set also includes an initial timestamp field. As the $decided$ set is empty, such timestamps are not changed in neither $p$ or $p'$ (l. \ref{algline:conditionchangets}). Therefore, both processes consider the same timestamp value for every message. Besides, as the $msgSet$ is the same for both, the $decided$ set remains identical for both $p$ and $p'$.\\

\noindent Induction step: Suppose that $p$ and $p'$ have already learnt the decisions of instance $k$, their $decided$ sets remained identical and they decided the same timestamp value of each message sent from some process in their group so far. As we assume that all processes within a group learn the decisions in the same consensus instace identifier order, then both $p$ and $p'$ will next learn the decision of the same instance $k+1$. From the agreement property of consensus, the $msgSet$ decided is the same for both processes. In the Algorithm \ref{algorithm:deliveryminimal}, the timestamps may be changed after the consensus decision only in line \ref{algline:changetsafterdecision}, based on what is already in the $decided$ set (l. \ref{algline:conditionchangets}). As the $msgSet$ and $decided$ sets are each identical in $p$ and $p'$, they will make the exact same change to the timestamp of each message in the \textit{while} loop (l. \ref{algline:whiledecided} to l. \ref{algline:rmothers}). Therefore, they also agree on the timestamps of each message decided on consensus instance $k+1$ and their $decided$ sets remain identical.\\

Now, consider the case of $p'$ not belonging to the same group of $p$. Let $q$ be any process of $group(p)$. After setting the final timestamp of $m$ (l. \ref{algline:whiledecided} to l. \ref{algline:rmothers}), $q$ \rmc{}s $m$ to $group(p')$ (l. \ref{algline:rmothers}). Once $p'$ \rmd{}s $m$, it never changes the timestamp of $m$, which already contains the final value, set by $q$ in accordance with $p$. As any two processes on the same group from which $m$ has been multicast agree on its timestamp, and since any process from other group which is also a destination of $m$ agree on its timestamp as well, than any two processes that are destinations of $m$ agree on its timestamp.
\end{proof}

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:respecttimestamps}
If two messages $m$ and $m'$ have both a correct process $p$ as a destination, and $m.ts < m'.ts$, then $p$ delivers $m'$ after delivering $m$.
\end{lems}

\begin{proof}
Consider Algorithm \ref{algorithm:deliveryminimal}. Suppose, by way of contradiction, that $m'$ has already been delivered by $p$, whereas $m$ has not. Message $m$ might have already been inserted into $barPending$ or not. In the former case, $m'$ could not have been delivered, since $m.ts < m'.ts$, thus it would not satisfy the condition from line \ref{algline:inbarpending} until $m$ has been delivered, so we have a contradiction in the case that $m$ was already in $barPending$.

The other case is if $m$ has not been inserted into $barPending$. From line \ref{algline:inbarpending}, we know that $p$ has received some barrier $b$ from every group in $sendersTo(group(p))$. Therefore, from Lemma \ref{lemma:oncedecidednotsmaller}, and since a barrier is also a message, we know that any new message that arrives at $p$ will have a timestamp greater than $m'.ts$. As $m$ has arrived after $m'$, then $m.ts > m'.ts$, which is also a contradiction.

We have proven that the timestamp order will not be violated by $p$. Considering that, once a message has been multicast, it will be delivered (Proposition \ref{props:validity}), then we know that all messages will be delivered by $p$, and in the correct timestamp order.\end{proof}

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props} \label{props:atomicorder}
If processes $p$ and $p'$ are both in $m.dst$ and $m'.dst$, then $p$ delivers $m$ before $m'$ if, and only if, $p'$ delivers $m$ before $m'$.
\end{props}

\begin{proof}
Immediate from Lemma \ref{lemma:agreetimestamps}, Lemma \ref{lemma:respecttimestamps} and Proposition \ref{props:agreement}.
\end{proof}

@@

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:fifofinalts}
If a process $p$ \cms{}s $m$, then $m'$, then the final values of their timestamps will be such that $m.ts < m'.ts$.
\end{lems}

\begin{proof}
From lines \ref{algline:gettime} and \ref{algline:rmlocal}, and the properties of the \rmc{} primitive, we know that all correct processes \rmd{} $m$ and then $m'$, which means that each process $p'$ from $group(p)$ also inserts them into its $propPending$ list in this same order.

Suppose, by way of contradiction, that $m'$ is decided before $m$. This implies that some process $p$ proposed a message set that included $m'$, but not $m$. As $p$ inserted $m$ before $m'$ in its $propPending$ list, it means that $m$ was removed by $p$ from such list before being included in some proposed message set. However, the only possibility of removing a message from $propPending$ is on line \ref{algline:removedecided}, which means that $m$ already belonged to the $decided$ set, so it had been decided before $m'$, which is a contradiction. Therefore, $m'$ is not decided before $m$.

This also means that $m'$ is not inserted into the $decided$ list before $m$ has already been inserted. From line \ref{algline:gettime}, we know that the initial timestamp of $m$ is already smaller than that of $m'$. The only way to change this order is by making the final timestamp $m.ts$ bigger than $m'.ts$ on line \ref{algline:changetsafterdecision}. Even if this happens, as $m'$ is inserted into $decided$ after $m$, from line \ref{algline:conditionchangets}, we know that the final timestamp of $m'$ will be made greater than that of $m$.
\end{proof}

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props} \label{props:fifoorder}
If $p$ \cms{}s $m$ and then $m'$, then no process $p'$ in both $m.dst$ and $m'.dst$ delivers $m'$ before delivering $m$.
\end{props}

\begin{proof}
Immediate from Lemma \ref{lemma:respecttimestamps} and Lemma \ref{lemma:fifofinalts}.
\end{proof}
